---
title: "R coding Exercise"
output: 
  html_document:
    toc: FALSE
---


#Load packages, view data
```{r}
#load dslabs and tidyverse packages
library("dslabs")
library("tidyverse") 
#look at gapminder help file
help(gapminder)
#get data structure overview
str(gapminder)
#get data summary
summary(gapminder)
#determine gapminder object type
class(gapminder)
```

#Create new objects as subset of original dataset
```{r}
#Assign African countries to new object
africadata <- gapminder %>% filter(continent=="Africa")
#get data structure overview for africadata
str(africadata)
#get data summary
summary(africadata)

#make new variable containing infant_mortality and life_expectancy
mort_life <- africadata %>% select(infant_mortality, life_expectancy)

#make new variable containing population and life_expectancy
pop_life <- africadata %>% select(population, life_expectancy)

#view data structure for new variables
str(mort_life)
str(pop_life)

#view data summaries for new variables
summary(mort_life)
summary(pop_life)
```

#Create plots of selected data
```{r}
#plot life expectancy vs infant mortality
ggplot(mort_life, aes(infant_mortality, life_expectancy)) + geom_point()
#In this plot we see a negative correlation between infant mortality and life expectancy
#There were also 226 missing values exluded from this plot

#plot life expectancy vs population (log scale)
ggplot(pop_life, aes(population, life_expectancy)) + geom_point() +scale_x_log10()

```
In this plot, we see a positive correlation between population and life expectancy, in "streaks". Looking at the original dataset, africadata, shows that a wide span of yeares (1960 to 2016) is probably the source of the feature in the plot. Another note: there were 51 observations with missing values for this plot. 

#Data cleaning
```{r}
#Let's select one year with a large amount of data to eliminate the limitations we saw above. 

#Figure out which years have missing infant mortality data


africadata[!complete.cases(africadata$infant_mortality),] #this line of code will show us the opposite of complete cases (meaning missing or NA cases) for the infant_mortality variable within our dataset
#we can see that infant_mortality is missing from 1960 to 1981, and for the year 2016

#We'll look at the year 2000 instead

#extract observations for year 2000 only
africadata_2000 <- africadata %>% filter(year==2000)

#check data structure for new object
str(africadata_2000)

#check summary of data for new object
summary(africadata_2000)
```

#Recreate plots
```{r}
#recreate above plots for year 2000 only
#plot life expectancy vs infant mortality
ggplot(africadata_2000, aes(infant_mortality, life_expectancy)) + geom_point()
#we see a negative correlation between infant mortality and life expectancy in this plot

#plot life expectancy vs population (log scale)
ggplot(africadata_2000, aes(population, life_expectancy)) + geom_point() +scale_x_log10()
#we don't see any obvious relationship between population and life expectancy in this plot
```

#Apply statistics to data
```{r}
#use a linear model to fit the data with infanct mortality as the predictor
fit1 <- lm(life_expectancy ~ infant_mortality, data = africadata_2000)

#use a linear model with population size as the predictor
fit2 <- lm(life_expectancy ~ population, data = africadata_2000)

#view summary of fit1 and fit2
summary(fit1)
summary(fit2)
```

Conclusion:
The p-value for the linear model using infant mortality as the predictor is < 0.05, while the p-value for the linear model using population as the predictor is > 0.05. This suggests that infant mortality correlates significantly with life expectancy and is a better predictor of life expectancy than is population size. 


